How is experience transformed into memory?

### 2018-08-28 paper/main.tex
added paxton to readme and paper
saving abstract and intro edits

\title{How is experience transformed into memory?}
% \title{Capturing the transformation of experience into memory}
\author{Andrew C. Heusser and Jeremy R. Manning\\Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755, USA\\Corresponding author: jeremy.r.manning@dartmouth.edu}

@@

\begin{abstract}
The human memory system is adept at cataloging the rich dynamics of ongoing experience. A defining feature of our everyday experiences is that, despite the rich complexities of real-world experiences, they unfold in a structured and often predictable manner. This structure gives each of our experiences (and thus our memories for those experiences) a unique `shape' with respect to how it unfolds in the relevant representational space.  However, temporal structure is typically absent (or severely impoverished) in traditional trial-based  experiments. As a consequence, our understanding of how the temporal structure (i.e., the shape) of our experiences influences memorability and memory organization is limited. In this study, we investigate how people verbally recount a video by characterizing and relating the thematic dynamics, or ``trajectories,'' of the video to participants' later recounting of the narrative. Using a novel approach that leverages topic modeling and Hidden Markov Models, we found that, despite substantial variability in the quantity of information recalled across participants as well as the particular words used to describe the experience, the overall shape of the stimulus was recapitulated by nearly every participant. Our findings suggest that assessing the degree of match between the shape of an experience and the subsequent retelling of that experience provides a much more nuanced view (compared with traditional methods, such as comparing the proportion of remembered details) into what it means to remember. These findings provide a window into which aspects of naturalistic experiences must be preserved, and which might be more flexible, in considering whether and how those experiences are remembered.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a binary operation: either a studied item is recalled or it isn't. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' an experience or a general feeling of ``familiarity'' \citep{Yone02}. However, characterizing and evaluating memory in more realistic contexts (e.g., recounting a recent experience to a friend) concerns a much different set of criteria.  First, eal-world recall is continuous, rather than binary.  Second, the specific words used to describe an experience have very little bearing on whether the experience is considered to have been ``remembered.''  Third, one might remember the gist of an experience but forget (or neglect to recount) particular details.  Or different people who share an experience might recount the experience with a similar level of detail, but the specific details that were remembered might vary across people.  Which aspects of those recollections should be considered fundamental to determining whether or how reliably an experience has been remembered?

@@

Human memory studies typically assess memory for the elements within an experience, such as the people, places and things encountered. However, real life experiences can also be described in terms of their temporal structure (i.e., how the contents relate/change over time).  For instance, naturalistic experiences are autocorrelated in space and time on short timescales. In other words, the contents of an experience are often similar from moment to moment and change gradually. As an example, many of us spend our mornings/evenings at home and much of our days at work (e.g.\ an office). While the specific happenings within these contexts differ, they are often highly correlated (and thus predictable) over time. Furthermore, our experiences can also be characterized by longer timescale correlations. For example, consider returning to your office after taking a lunch break. The spatial contexts before and after lunch are highly similar, but separated in time. Thus, while evaluating the ability to recall the isolated elements of an experience has been a fruitful way to learn about human memory, another important dimension to consider is the temporal structure of the remembered experience, and how the temporal structure of the memory relates to the original experience. However, these recurrent and gradually changing temporal dynamics are not typically present in traditional memory studies, but are likely critical if we wish to understand how our memory systems remember our life experiences.

While our experiences often change gradually, they can also change rapidly (e.g.\ walking through a doorway) \citep{RadvZack17}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize our memories into ``events'' \citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}. Short timescale autocorrelations in the contents of an experience are thought associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, EzzyDava11, DuBrDava13}.  In other words, the interplay between stable and transient temporal dynamics may impose structure on memories for those experiences. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people verbally recount their experiences in a way that is structured around events has not been established.

In this study, we analyzed an open dataset in which participants view and then verbally recall an episode of the BBC series \textit{Sherlock} \citep{ChenEtal17}. To capture the temporal structure of the video, we developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17}. This model casts naturalistic stimuli as high-dimensional topic ``trajectories'', whereby the theme of each moment of the video is represented as a topic vector. In this framework, successful remembering entails verbally ``traversing'' the path formed by the stimulus topic trajectory as closely as possible. We hypothesized that successful memory for a naturalistic stimulus is given by a recall trajectory that closely follows the path (and covers the entirety) of the stimulus trajectory.

>>>>

\title{How is experience transformed into memory?}
% \title{Observing the transformation of dynamic experiences into memory}
\author{Andrew C. Heusser, Paxton C. Fitzpatrick \& Jeremy R. Manning\\Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755, USA\\Corresponding author: jeremy.r.manning@dartmouth.edu}

@@

\begin{abstract}
The human memory system is adept at cataloging the rich dynamics of ongoing experience. A defining feature of our everyday experiences is that, despite the complexities of real-world experiences, they unfold in a structured and often predictable manner. This structure gives each of our experiences (and thus our memories for those experiences) a unique `shape' with respect to how it unfolds in the relevant representational space.  However, temporal structure is typically absent (or severely impoverished) in traditional trial-based  experiments. As a consequence, our understanding of how the temporal structure (i.e., the shape) of our experiences influences memorability and memory organization is limited. In this study, we investigate how people verbally recount a video by characterizing and relating the thematic dynamics, or ``trajectories,'' of the video to participants' later recountings of the narrative. Using a novel approach that leverages topic modeling and Hidden Markov Models, we found that, despite substantial variability in the quantity of information recalled across participants as well as the particular words used to describe the experience, the overall shape of the stimulus was recapitulated by nearly every participant. Our findings suggest that assessing the degree of match between the shape of an experience and the subsequent retelling of that experience provides a much more nuanced view (compared with traditional methods such as comparing the proportion of remembered details) into what it means to remember. These findings provide a window into which aspects of naturalistic experiences must be preserved, and which might be more flexible, in considering whether and how those experiences are remembered.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a binary operation: either a studied item is recalled or it isn't. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' an experience or a general feeling of ``familiarity'' \citep{Yone02}. However, characterizing and evaluating memory in more realistic contexts (e.g., recounting a recent experience to a friend) concerns a much different set of criteria.  First, real-world recall is continuous, rather than binary.  Second, the specific words used to describe an experience have very little bearing on whether the experience is considered to have been ``remembered.''  Third, one might remember the gist or essence of an experience but forget (or neglect to recount) particular details.  Or different people who share an experience might recount the experience with a similar level of detail, but the specific details that were remembered might vary across people.  Which aspects of those recollections should be considered fundamental to determining to whast extent or how reliably an experience has been remembered?

@@

Human memory studies typically assess memory for the elements within an experience, such as the people, places and things encountered. However, real life experiences can also be described in terms of their temporal structure (i.e., how the contents relate/change over time).  For instance, naturalistic experiences are autocorrelated in space and time on short timescales. In other words, the contents of an experience often change gradually and are therefore similar from moment to moment. As an example, many of us spend our mornings/evenings at home and much of our days at work (e.g.\ in an office). While the specific happenings within these contexts differ, they are often highly correlated (and thus predictable) within an experience. Furthermore, our experiences can also be characterized by longer timescale correlations. For example, consider returning to your office after taking a lunch break. The spatial contexts before and after lunch are highly similar, but separated in time. Thus, while evaluating the ability to recall the isolated elements of an experience has been a fruitful way to learn about human memory, another important dimension to consider is the temporal structure of the remembered experience, and how the temporal structure of the memory relates to the original experience. These recurrent and gradually changing temporal dynamics are not typically present in traditional memory studies, but are likely critical to understanding how our memory systems encode, store and retrieve our life experiences.

While our experiences often change gradually, they can also sometimes change suddenly (e.g.\ walking through a doorway) \citep{RadvZack17}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize our memories into ``events'' \citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}. Short-timescale autocorrelations in the contents of an experience are thought to associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, EzzyDava11, DuBrDava13}.  In other words, the interplay between stable and transient temporal dynamics may impose structure on memories for those experiences. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people verbally recount their experiences in a way that is structured around events has not been established.

In this study, we analyzed an open dataset in which participants view and then verbally recall an episode of the BBC series \textit{Sherlock} \citep{ChenEtal17}. To capture the temporal structure of the video, we developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17}. This technique casts naturalistic stimuli as high-dimensional topic ``trajectories'', whereby the theme of each moment of the video is represented as a topic vector. In this framework, successful remembering entails verbally ``traversing'' the path formed by the stimulus topic trajectory as closely as possible. We hypothesized that successful memory for a naturalistic stimulus is given by a recall trajectory that closely follows the path (and covers the entirety) of the stimulus trajectory.
